Approximation as Differentiation

### 2025-07-10 main/approx-as-tangents.tex
laird07, fix ref.
[ tex ] more discussion of semantics GPS in Fam(LatGal)
[ tex ] Fix up some references and citations

For the purposes of \GPS, we are interested in using
approximations not to model computation on infinite objects, but instead
for revealing how programs explore their inputs when producing parts
of their output. Therefore, we ignore completeness properties of the
partially ordered sets we consider\footnote{\bob{we might want to bring this back when considering general recursive programs. Are bidomains \cite{berry79} useful? they have two orderings. See also Laird's Bistable Biorders.}}.

@@

When giving a denotational semantics for sequential programming
languages, Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions that whose
output cannot be assigned a unique minimal input that accounts for
it. Therefore, continuous functions are incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

@@

\begin{example}
  \leavevmode\\[-3ex]
  \begin{enumerate}
  \item The function $\mathrm{strictOr}$ is stable. For example, for
    the input-output pair
    $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input
    that gives this output is exactly $(\mathsf{tt}, \mathsf{ff})$. If
    we take the approximation $\bot \leq \mathsf{tt}$ of the output,
    then the corresponding minimal input is $(\bot, \bot)$. The
    function $\mathrm{shortCircuitOR}$ is also stable. For the
    input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$,
    the minimal input that gives this input is $(\mathsf{tt},\bot)$,
    indicating that the presence of $\mathsf{ff}$ in the second
    argument was not necessary to produce this output. As with
    $\mathrm{strictOr}$, the minimal input required to produce the
    output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

@@

As well as codifying exactly what we mean by differentiable structure
on partially ordered sets, proving this conjecture would also tell us
what higher derivatives mean in this context as well, something that
we have not considered above. We will extend this conjecture to our
refined model of lattice approximated sets in \secref{model-summary}.

>>>>

For the purposes of \GPS, we are interested in using
approximations not to model computation on infinite objects, but instead
for revealing how programs explore their inputs when producing parts
of their output. Therefore, we ignore completeness properties of the
partially ordered sets we consider.

@@

When giving a denotational semantics for sequential programming
languages, Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions whose
output cannot be assigned a unique minimal input that accounts for
it. Therefore, continuous functions are incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

@@

\begin{example}\label{ex:stable-functions}
  \leavevmode\\[-3ex]
  \begin{enumerate}
  \item The function $\mathrm{strictOr}$ is stable. For example, for
    the input-output pair
    $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input
    that gives this output is exactly $(\mathsf{tt}, \mathsf{ff})$. If
    we take the approximation $\bot \leq \mathsf{tt}$ of the output,
    then the corresponding minimal input is $(\bot, \bot)$. The
    function $\mathrm{shortCircuitOR}$ is also stable. For the
    input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$,
    the minimal input that gives this input is $(\mathsf{tt},\bot)$,
    indicating that the presence of $\mathsf{ff}$ in the second
    argument was not necessary to produce this output. As with
    $\mathrm{strictOr}$, the minimal input required to produce the
    output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

@@

As well as codifying exactly what we mean by differentiable structure
on partially ordered sets, proving this conjecture would also tell us
what higher derivatives mean in this context as well, something that
we have not considered above. We will extend this conjecture to our
refined model of lattice approximated sets in \secref{models:summary}.

### 2025-07-10 main/conclusion.tex
[ tex ] Fix up some references and citations

@@

Lists are the only recursive data type we provided in our source language, so important future work is
supporting general inductive and coinductive types. \citet{nunes2023} support automatic differentiation for
datatypes defined as the least or greatest fixed points of $\mu\nu$-polynomial functors; we could potentially
adopt a similar approach. Full inductive types would allow us to embed an interpreter for untyped lambda
calculus; combined with the CBN monadic translation described in \secref{todo} which uniformly inserts
approximation points, we should be able to obtain the program slicing behaviour of earlier \GPS work ``for
free''. Coinductive types (e.g.~streams) present additional challenges, especially for defining
join-preserving backward maps, but also open the door to slicing (finite prefixes of) infinite data sources,
with some likely relationship to the problem of dealing with partial or non-terminating computations.

>>>>

Are bidomains \cite{berry79} useful? they have two orderings. See also
Laird's Bistable Biorders \cite{laird07}.

@@

Lists are the only recursive data type we provided in our source language, so important future work is
supporting general inductive and coinductive types. \citet{nunes2023} support automatic differentiation for
datatypes defined as the least or greatest fixed points of $\mu\nu$-polynomial functors; we could potentially
adopt a similar approach. Full inductive types would allow us to embed an interpreter for untyped lambda
calculus; combined with the CBN monadic translation described in \secref{cbn-translation} which uniformly inserts
approximation points, we should be able to obtain the program slicing behaviour of earlier \GPS work ``for
free''. Coinductive types (e.g.~streams) present additional challenges, especially for defining
join-preserving backward maps, but also open the door to slicing (finite prefixes of) infinite data sources,
with some likely relationship to the problem of dealing with partial or non-terminating computations.

### 2025-07-10 main/language.tex
[ tex ] Fix up some references and citations

Assuming that $\cat{C}$ is bicartesian closed and has a list object (\propref{lists}), then we can extend an
interpretation of a signature $\Sigma$ to an interpretation of the whole language over
$\Sigma$. \figref{semantics:types} and \figref{semantics:contexts} define the interpretation of types and
contexts as objects of $\cat{C}$ respectively. Terms are interpreted as morphisms between the interpretations
of the context and type, as defined in \figref{semantics:terms}. We have used the notations $\pi_i$ for
projections, $\prodM{f}{g}$ for pairing, $\coprodM{f}{g}$ for (parameterised) copairing, $!_x$ for morphisms
to the terminal object, and $\lambda$ and $\eval$ for currying and evaluation for exponentials.

@@

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the query from \exref{introduction-example}.
\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. The approximation lattices for all types are
  always the trivial one-point lattice, and the forward and reverse approximation maps are necessarily the
  identity.
\item To get interesting approximation information, we take
  $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting-monad}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and $\sem{\mathrm{add}}_{\PrimOp}$ defined from
  the unlifted interpretations above and the monad structure. This interpretation recovers the behaviour given
  in \exref{introduction-example}: running the reverse approximation map at the input $\mathsf{a}$ with the
  top approximation maps indeed reveals that only the numbers in the rows tagged with that label in the input
  are used, and likewise for $\mathsf{b}$.
\item More quantitative approximation information can be obtained by using the interval approximation
  interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}.

  Recall that $\mathrm{query}\,\mathsf{a}\,\mathit{db} = 1$, so in the reverse direction we can pick an interval containing $1$ to discover the largest (i.e. least in the order) intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we pick $[-0.9,1]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query}\,\mathsf{a}\,\mathsf{db})_r([0.9,1.1]) = [(\cdot,[-0.1,0.1]),(\cdot,\bot),(\cdot,[0.9,1.1])]
  \end{displaymath}
  Indicating that, to achieve an output within $[0.9,1.1]$ for this query, either the first $\mathsf{a}$ row
  could be in $[-0.1,0.1]$ or the second one could be in $[0.9,1.1]$.
\end{enumerate}

@@

We can now carry out the systematic insertion of approximation points that we foreshadowed in
\secref{models-of-total-gps:lifting-monad}, using \citet[\S 3.1]{notions-of-computation}'s monadic CBN translation. We use the $T$
monad from \secref{tagging-monad} because it can be defined in terms of the language constructed we already
have. This requires that we have a signature $\Sigma$ that includes a primitive type to be interpreted as the
approximation object $\mathbb{A}$ and primitive operations to be interpreted as the monoid operations on this
object.

>>>>

Assuming that $\cat{C}$ is bicartesian closed and has a list object (\eqnref{lists}), then we can extend an
interpretation of a signature $\Sigma$ to an interpretation of the whole language over
$\Sigma$. \figref{semantics:types} and \figref{semantics:contexts} define the interpretation of types and
contexts as objects of $\cat{C}$ respectively. Terms are interpreted as morphisms between the interpretations
of the context and type, as defined in \figref{semantics:terms}. We have used the notations $\pi_i$ for
projections, $\prodM{f}{g}$ for pairing, $\coprodM{f}{g}$ for (parameterised) copairing, $!_x$ for morphisms
to the terminal object, and $\lambda$ and $\eval$ for currying and evaluation for exponentials.

@@

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the query from \exref{introduction-example}.
\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. The approximation lattices for all types are
  always the trivial one-point lattice, and the forward and reverse approximation maps are necessarily the
  identity.
\item To get interesting approximation information, we take
  $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. This
  interpretation recovers the behaviour given in \exref{introduction-example}: running the reverse
  approximation map at the input $\mathsf{a}$ with the top approximation maps indeed reveals that only the
  numbers in the rows tagged with that label in the input are used, and likewise for $\mathsf{b}$.
\item More quantitative approximation information can be obtained by using the interval approximation
  interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}.

  Recall that $\mathrm{query}\,\mathsf{a}\,\mathit{db} = 1$, so in the reverse direction we can pick an
  interval containing $1$ to discover the largest (i.e. least in the order) intervals that will give rise to
  this output as \emph{independent} changes to the input. For example, if we pick
  $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query}\,\mathsf{a}\,\mathsf{db})_r([\frac{9}{10},\frac{11}{10}]) = [(\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}])]
  \end{displaymath}
  Indicating that, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$ for this query, either the first
  $\mathsf{a}$ row could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in
  $[\frac{9}{10},\frac{11}{10}]$.
\end{enumerate}

@@

We can now carry out the systematic insertion of approximation points that we foreshadowed in
\secref{models-of-total-gps:lifting}, using \citet[\S 3.1]{notions-of-computation}'s monadic CBN
translation. We use the $T$ monad from \secref{tagging-monad} because it can be defined in terms of the
language constructed we already have. This requires that we have a signature $\Sigma$ that includes a
primitive type to be interpreted as the approximation object $\mathbb{A}$ and primitive operations to be
interpreted as the monoid operations on this object.

### 2025-07-10 main/models.tex
[ tex ] Fix up some references and citations

Using the infinitary coproducts and finite products, we can construct
a wide range of other useful semantic models of datatypes in
$\Fam(\cat{C})$. For example, lists can be constructed as a coproduct
\begin{displaymath}
  \List(X) = \coprod_{n \in \mathbb{N}} X^n
\end{displaymath}
where $X^0 = 1$ (the terminal object) and $X^{n+1} = X \times X^n$.

@@

With this embedding functor, we see in
\propref{first-order-type-interp} that the interpretation of
first-order types will be the same up to isomorphism in
$\Fam(\LatGal)$ and $\Fam(\MeetSLat \times \JoinSLat^\op)$, as long as
we interpret the base types as objects in $\Fam(\LatGal)$. At
higher-order, however, the meet-semilattice and join-semilattice sides
of the interpretation will diverge, and it is no longer clear that the
interpretation of programs using higher-order functions internally
will result in Galois connections. In \secref{definability} we will
see that every program with first-order type (even if it uses
higher-order functions internally) does in fact have an interpretation
definable in $\Fam(\LatGal)$.

@@

\begin{remark}
  \label{rem:further-structure}
  These examples highlight a potential criticism of $\Fam(\LatGal)$ as
  a category for modelling \GPS. For $\mathrm{shortCircuitOr}$, we had
  $(\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top)
  = (\top, \bot)$, indicating that the second argument was not needed
  for computing the output. However, there is no way, in
  $\Fam(\LatGal)$, of turning this into a rigorous statement that the
  $\Set$-component of this morphism does not actually depend on its
  second argument. We conjecture that this can be rectified by
  requiring some kind of additional structure on each object
  $(X, \partial X)$ consisting of a map
  $\Pi_{x : X}.\partial X(x) \to \mathcal{P}(X)$, where $\mathcal{P}$
  is the powerset, which identifies for each $x$ which elements are
  indistinguishable from $x$ at this level of approximation. One would
  also presumably have to require additional conditions for this to
  respect the lattice structure and be preserved by
  morphisms\footnote{This additional structure is reminiscent of the
    additional structure on \emph{directed containers} defined by
    \citet{ahman-uustalu}. They require a map
    $\Pi_{x : X}.\partial X(x) \to X$ picking out a specific $X$
    ``jumped to'' by some change $\delta x : \partial X(x)$ at $x$. In
    our proposed setup, we follow the approximation theme of \GPS by
    having a \emph{set} of things that could be taken instead of the
    original $x$. Objects of $\Fam(\LatGal)$ are ``directed'' in this
    sense because the approximation map can pick out the element of
    the original poset that was approximating $x$.}.
\end{remark}

@@

\subsection{Summary}

>>>>

Using the infinitary coproducts and finite products, we can construct
a wide range of other useful semantic models of datatypes in
$\Fam(\cat{C})$. For example, lists can be constructed as a coproduct
\begin{equation}
  \label{eqn:lists}
  \List(X) = \coprod_{n \in \mathbb{N}} X^n
\end{equation}
where $X^0 = 1$ (the terminal object) and $X^{n+1} = X \times X^n$.

@@

With this embedding functor, we will see in
\lemref{first-order-agreement-types} that the interpretation of
first-order types will be the same up to isomorphism in
$\Fam(\LatGal)$ and $\Fam(\MeetSLat \times \JoinSLat^\op)$, as long as
we interpret the base types as objects in $\Fam(\LatGal)$. At
higher-order, however, the meet-semilattice and join-semilattice sides
of the interpretation will diverge, and it is no longer clear that the
interpretation of programs using higher-order functions internally
will result in Galois connections. In \secref{definability} we will
see that every program with first-order type (even if it uses
higher-order functions internally) does in fact have an interpretation
definable in $\Fam(\LatGal)$.

@@

\begin{remark}
  \label{rem:further-structure}
  These examples highlight a potential criticism of $\Fam(\LatGal)$ as
  a category for modelling \GPS. For $\mathrm{shortCircuitOr}$, we had
  $(\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top)
  = (\top, \bot)$, indicating that the second argument was not needed
  for computing the output. However, there is no way, in
  $\Fam(\LatGal)$, of turning this into a rigorous statement that the
  $\Set$-component of this morphism does not actually depend on its
  second argument. We conjecture that this can be rectified by
  requiring some kind of additional structure on each object
  $(X, \partial X)$ consisting of a map
  $\Pi_{x : X}.\partial X(x) \to \mathcal{P}(X)$, where $\mathcal{P}$
  is the powerset, which identifies for each $x$ which elements are
  indistinguishable from $x$ at this level of approximation. One would
  also presumably have to require additional conditions for this to
  respect the lattice structure and be preserved by
  morphisms\footnote{This additional structure is reminiscent of the
    additional structure on \emph{directed containers} defined by
    \citet{ahman-chapman-uustalu2012}. They require a map
    $\Pi_{x : X}.\partial X(x) \to X$ picking out a specific $X$
    ``jumped to'' by some change $\delta x : \partial X(x)$ at $x$. In
    our proposed setup, we follow the approximation theme of \GPS by
    having a \emph{set} of things that could be taken instead of the
    original $x$. Objects of $\Fam(\LatGal)$ are ``directed'' in this
    sense because the approximation map can pick out the element of
    the original poset that was approximating $x$.}.
\end{remark}

@@

\subsection{Summary}
\label{sec:models:summary}
